---
title: "Titanic_readme"
author: "Michal Rackiewicz"
date: "2023-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Load packages and data

```{r}
library(tidyverse)
library(Hmisc)
library(corrplot)
library(ggpol)
```

```{r titanic data}
train <- read_csv("./Data/train.csv", col_types = "nffcfnnncnf")
test <- read_csv("./Data/test.csv", col_types = "nffcfnnncnf")
```

### Exploratory data analysis

```{r}
str(train)
```

First we will look at the summary statistics for each variable

```{r}
summary(train)
```

Let us also check the distribution of the dependent variable. By doing this we get an idea of how balanced the data set is and whether we have to do something to balance groups before modeling.

```{r}
train %>% 
  select(Survived) %>% 
  group_by(Survived) %>% 
  summarise(n = n()) %>% 
  mutate(Freq = paste0(round(100 * n/sum(n), 1), "%"))
```

We also want to check for missing values.

```{r}
train %>% 
  summarise_all(list(~sum(is.na(.))))
```

We see that although "Age" is the only variable with missing values. We will replace blank cells and those containing only spaces, as well as zeros in Fare with NAs.

```{r}
train$Fare[train$Fare == 0] = NA
```

```{r}
train <- train %>% 
  mutate(across(everything(), ~gsub("^[[:space:]]*$", NA, .)))
```

```{r}
train %>% 
  summarise_all(list(~sum(is.na(.))))
```

Due to too many missing values, we will drop the Cabin column.

```{r}
train <- train %>% select(!Cabin)
```


## Data visualization
First, we will look at some basic plots of variable distributions and 

```{r numeric plots, echo=FALSE}
train %>% 
  select(where(is.numeric))%>%
  hist.data.frame(.)
```



```{r}
str(train)
```

```{r}
train <- train %>% 
  mutate(across(c(PassengerId, Age, SibSp, Parch, Fare), as.integer),
         across(c(Survived, Pclass, Embarked, Sex), as.factor)
         )
```

```{r}
pal_c <- wes_palette("GrandBudapest1", 21, type = "continuous")
```


Distributions of passenger IDs across passenger classes.

```{r}
train
ggplot(train, aes(x = PassengerId,
                  y = Pclass,
                  fill = Pclass)) +
  geom_jitter(shape = 21,
             color = "white") +
  labs(title = "Distirbution of passenger IDs",
       y = "Class",
       x = "Passenger ID",
       fill = "Passenger class") +
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```
Since the passenger IDs are distributed uniformly across all classes we will remove it as it holds no predictive value.


Let us also inspect the relationships between the variables. First, we will a plot a correlation map displaying linear relationship between the variables.
```{r}
correlation_matrix <- train %>%
  select(!c(Name, PassengerId, Ticket)) %>%
  mutate(across(where(is.factor), as.numeric)) %>%
  filter(complete.cases(.)) %>% 
  cor(as.matrix(.), method = "spearman")
```

```{r, echo=FALSE}
corrplot(correlation_matrix, 
         method = "square",
         type = "lower",
         order = "FPC",
         addCoef.col = 'black',
         diag = FALSE
         )
```
Correlation matrix with Age*Sex



Let us now take a closer look at those variables which are correlated with survival. As expected sex is a strong predictor of survival. Let us look at the numbers of survivors based on sex and passenger class.

```{r, echo=FALSE}
ggplot(train, 
       aes(x = interaction(Pclass, Sex),
           fill = Survived)) +
  geom_bar(stat="count",
           width = 0.4,
           position = position_dodge(width = 0.6)) +
  labs(title = "Survival based on Passenger Class and Sex",
       y = "Count",
       x = "Class/Sex",
       fill = "Survived") +
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```

If we look at the distribution of age based on survival we can see that 
```{r}
ggplot(train,
       aes(x = Age)) +
  geom_histogram(aes(fill = Survived,
                     alpha = 0.5)) +
  labs(title = "Age distribution of men in each class",
     y = "Age",
     x = "Class",
     fill = "Survived") +
  facet_grid() +
  theme_minimal() +
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```



```{r}
train %>% 
  filter(Sex == "male") %>%
  ggplot(aes(x = Pclass, 
             y = Age))+
  geom_boxjitter(aes(fill = Survived),
                 position = position_dodge(width = 0.8),
                 width = 0.3,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4)+
  labs(title = "Age distribution of male passengers in each class",
       y = "Age",
       x = "Class",
       fill = "Survived")+
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```

```{r}
train %>% 
  filter(Sex == "female") %>%
  ggplot(aes(x = Pclass, 
             y = Age))+
  geom_boxjitter(aes(fill = Survived),
                 position = position_dodge(width = 0.8),
                 width = 0.3,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4)+
  labs(title = "Age distribution of female passengers in each class",
       y = "Age",
       x = "Class",
       fill = "Survived")+
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```

```{r}
train %>% 
  filter(Survived == 1) %>%
  ggplot(aes(x = Pclass, 
             y = Age))+
  geom_boxjitter(aes(fill = Sex),
                 position = position_dodge(width = 0.8),
                 width = 0.3,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4)+
  labs(title = "Age of survivors based on Class and Sex",
       y = "Age",
       x = "Class",
       fill = "Sex")+
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```

Passenger class and fare are strongly negatively correlated, i.e. the highest fares are found in class 1.

```{r, echo=FALSE}
ggplot(train, aes(x = Pclass,
                  y = Fare,
                  fill = Pclass)) +
  geom_boxplot() +
  labs(title = "Fares per passenger class",
       y = "Fare",
       x = "Class") +
  theme_minimal()+
  scale_fill_manual(values = wes_palette("GrandBudapest1"))
```

```{r}
train %>% 
  group_by(Survived) %>% 
  summarise(across(c(Pclass, Sex, SibSp, Parch, Embarked), ~n()))
  
```


