---
title: "Titanic"
author: "Michal Rackiewicz"
date: "2023-09-29"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages and data

```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(skimr)
library(RColorBrewer)
library(viridis)
library(Hmisc)
library(corrplot)
library(ggpubr)
library(ggpol)
library(gridExtra)
library(rlang)
library(knitr)
library(scales)
```

Set options
```{r options}
options(digits = 2,
        # do not show warning messages
        warn = -1,
        # do not user scientific number formatting
        scipen = 999)

theme_set(theme_minimal())
```

Load data
```{r load_data, warning=FALSE}
train <- read_csv("./Data/train.csv", col_types = "nffcfnffcncf")
test <- read_csv("./Data/test.csv", col_types = "nffcfnffcncf")
```

## Exploratory data analysis
```{r}
str(train)
```

```{r summary_stats}
skimr::skim(train)
```

### Missing values per passenger
```{r}
train %>% 
  mutate(across(everything(), as.character)) %>% 
  pivot_longer(cols = !Name, names_to = "cat", values_to = "val") %>% 
  filter(is.na(val)) %>% 
  count(Name, sort = TRUE)
```

```{r class_balance}
train %>% 
  select(Survived) %>% 
  group_by(Survived) %>% 
  summarise(Count = n()) %>% 
  mutate(Proportion = paste0(round(100 * Count/sum(Count), 1), "%")) 
```

```{r fare_NA, echo=TRUE}
train <- train %>%
  # encode blank entries and spaces
  mutate(
    across(
      everything(),
      ~if_else(. %in% c("", " "), NA, .)
    ),
    # encode 0 values in fare
    Fare = if_else(Fare == 0, NA, Fare),
    across(c(SibSp, Parch, Pclass), fct_inseq)
  )
```

## Base model recipe
```{r base_recipe}
base_rec <- recipe(Survived ~ ., train)
summary(base_rec)
```

```{r cv_folds}
# create folds for cross-validation
set.seed(123)
train_folds <- vfold_cv(data = train, repeats = 2, strata = Survived)

train_folds
```


### Numeric variables
```{r}
plot_vars <- function(variable_type) {
  
  vars <- train %>% 
    select(where(paste0("is.", as_string(variable_type)))) %>% 
    names()
  
  geom <- NULL
  
  fill_scale <- NULL
  
  plots <- list()
  
  if (variable_type == "double") {
    geom <-  geom_density()
  } else {
    geom <-  geom_bar()
    fill_scale <- scale_fill_brewer(palette = "Dark2", type = "qual")   
  }

  for (i in seq_along(vars)) {
    
    var <- vars[i]
    
    p <- train %>% 
      filter(!is.na(!!rlang::sym(var))) %>% 
      ggplot(aes_string(x = var,
                        fill = var)) +
      geom +
      labs(title = paste("Distribution of", var),
         y = "Count",
         x = var) +
      theme(legend.position = "none") +
      fill_scale
    
    plots <- append(plots, list(p))
  }
  
  do.call(gridExtra::grid.arrange, c(plots, ncol = 2))
    
}
```

```{r}
plot_vars("numeric")
```

### Correlation map
```{r corr_matrix, include=FALSE}
correlation_matrix <- train %>%
  select(where(is.numeric)) %>%
  filter(complete.cases(.)) %>% 
  hmisc::cor(as.matrix(.), method = "spearman")
```

```{r corr_plot, echo=FALSE}
corrplot::corrplot(correlation_matrix, 
         method = "square",
         type = "lower",
         order = "alphabet",
         addCoef.col = 'black',
         diag = FALSE)
```

### Categorical variables
```{r categorical_vars, echo=FALSE, message=FALSE, warning=FALSE}
categorical_plots <- list()
cat_vars <- train %>% 
  select(where(is.factor)) %>% 
  names()

for (i in seq_along(cat_vars)) {
  
  var <- cat_vars[i]
  var_levels <- length(unique(train[[var]]))
  # colors_cat <- 
  
  p <- train %>% 
    filter(!is.na(!!rlang::sym(var))) %>% 
    ggplot(aes_string(x = var,
                       fill = var)) +
    geom_bar() +
    labs(title = paste("Distribution of", var),
       y = "Count",
       x = var) +
    scale_fill_brewer(palette = "Dark2", type = "qual") +
    theme(legend.position = "none")
  
  categorical_plots <- append(categorical_plots, list(p))
}

do.call(grid.arrange, c(categorical_plots, ncol = 2))
```
#### Chi-square test
```{r}
train %>% 
```



### Passenger ID
```{r id_survival_plot, echo=FALSE, message=FALSE, warning=FALSE}
train %>% 
  mutate(Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = PassengerId)) +
  geom_histogram(aes(fill = Survived,
                     bin = 50),
                 color = "black") +
  labs(title = "Distribution of passenger IDs",
     y = "Count",
     x = "Passenger ID",
     fill = "Survived") +
  facet_grid(Sex ~ .) +
  scale_fill_brewer(palette = "Dark2", type = "qual")
```

```{r id_pclass_plot, echo=FALSE}
train %>% 
  ggplot(aes(y = PassengerId,
             x = Pclass,
             fill = Pclass)) +
  geom_boxplot() +
  labs(title = "Distirbution of passenger IDs",
       y = "Passenger ID",
       x = "Passenger Class",
       fill = "Passenger class") +
  scale_fill_brewer(palette = "Dark2", type = "qual") +
  theme(legend.position = "none")

```

### Sex
```{r sex_survival_plot, echo=FALSE, message=FALSE, warning=FALSE}
train %>%
  group_by(Survived, Sex) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(percentage = round(n*100/sum(n), 2),
         label = paste0(percentage,"%"),
         Survived = case_match(Survived,
                               "0" ~ "deceased",
                               "1" ~ "survived")) %>% 
  ggplot(aes(x = Sex,
             y = n,
             fill = Sex,
             label = label)) +
  geom_col() +
  facet_grid(~Survived) +
  labs(title = "Survival ratios for each sex",
       y = "Passenger count",
       x = "Sex",
       fill = "Sex") +
  scale_fill_brewer(palette = "Dark2", type = "qual") +
  theme(legend.position = "none")
```
### Passenger class
```{r pclass_survival_plot, echo=FALSE, message=FALSE, warning=FALSE}
train %>% 
  group_by(Pclass, Survived) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  mutate(percentage = round(n*100/sum(n), 2), 
         label = paste0(percentage,"%"),
         Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = fct_inseq(Pclass),
             y = n,
             fill = Survived,
             label = label)) +
    geom_col() +
    labs(title = "Survival across classes",
         y = "Passenger count",
         x = "Passenger class",
         fill = "Survival status") +
  geom_label(position = position_stack(0.5),
             show.legend = FALSE) +
  scale_fill_brewer(palette = "Dark2", type = "qual")
```
### Fare

```{r echo=FALSE, warning=FALSE}
train%>% 
  ggplot(aes(x = fct_inseq(Pclass),
             y = Fare)) +
  geom_boxplot(aes(color = Pclass)) +
  labs(title = "Fares in each passenger class",
       y = "Fare",
       x = "Passenger class") +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Dark2", type = "qual")
```
```{r}
train %>% 
  group_by(fct_inseq(Pclass)) %>% 
  summarise(Fare_mean = mean(Fare),
            Fare_median = median(Fare),
            SD = sd(Fare))
```

```{r}
train %>% 
  group_by(SibSp) %>% 
  summarise(Fare_mean = mean(Fare),
            Fare_median = median(Fare),
            SD = sd(Fare))
```


```{r}
# impute mean based on Pclass
rec <- rec %>% 
  step_impute_bag(impute_with = Pclass)
```



```{r}
train %>% mutate(across(c(Pclass, SibSp, Parch), fct_inseq))
```



### Age
```{r age_sex_survival_histogram,echo=FALSE, message=FALSE, warning=FALSE}
train %>% 
  mutate(Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = Age)) +
  geom_histogram(aes(fill = Survived,
                     bin = 50),
                 color = "black") +
  labs(title = "Age distribution of male and female passengers",
     y = "Count",
     x = "Age",
     fill = "Survival status") +
  facet_grid(Sex ~ .) +
  scale_fill_brewer(palette = "Dark2", type = "qual")
```
#### Age distribution for each sex
```{r}
train %>% 
  filter(!is.na(Age)) %>% 
  ggplot(aes(Sex, Age)) +
  geom_boxplot(aes(color = Sex)) +
  scale_color_brewer(palette = "Dark2", type = "qual")
```

#### Age distribution for both sexes in each passenger class
```{r age_sex_survival_boxplot, echo=FALSE, warning=FALSE}
train %>% 
  mutate(Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = fct_inseq(Pclass), 
             y = Age)) +
  ggpol::geom_boxjitter(aes(fill = Survived),
                 position = position_dodge(width = 0.5),
                 width = 0.4,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4) +
  labs(title = "Age distribution of male passengers in each class",
       y = "Age",
       x = "Passenger class",
       fill = "Survival status") +
  facet_grid(~Sex) + 
  RColorBrewer::scale_fill_brewer(palette = "Dark2", type = "qual")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
train %>%
  group_by(Pclass, Sex, Survived) %>% 
  summarise(n = n(), .group = "drop") %>%  
  mutate(Percent = round(n*100/sum(n)),
         Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = interaction(fct_inseq(Pclass), Sex),
             y = Percent,
           fill = Survived)) +
  geom_bar(stat="identity",
           width = 0.4,
           position = position_dodge(width = 0.6)) +
  labs(title = "Survival based on Passenger Class and Sex",
       y = "Passengers [%]",
       x = "Class.Sex",
       fill = "Survival status") +
  RColorBrewer::scale_fill_brewer(palette = "Dark2", type = "qual")
```


### Embarked
```{r embarked_survival_barplot, echo=FALSE, message=FALSE}
train %>% 
  filter(!is.na(Embarked)) %>% 
  group_by(Embarked, Survived) %>% 
  summarise(n = n(), .groupes = "drop") %>% 
  mutate(percent = round(n*100/sum(n), 2), 
         label = paste0(percent,"%"),
         Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  ggplot(aes(x = Embarked,
             y = n,
             fill = Survived,
             label = label)) +
    geom_col() +
    labs(title = "Numbers of survivors per port of embarkation",
         y = "Count",
         x = "Port of embarkation",
         fill = "Survival status") +
  geom_label(position = position_stack(0.5),
             show.legend = FALSE) +
  scale_fill_brewer(palette = "Dark2", type = "qual")
  
```

```{r embarked_pclass_barplot, echo=FALSE, message=FALSE}
train %>% 
  filter(!is.na(Embarked)) %>%
  group_by(Embarked, Pclass) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(prop = count/sum(count)) %>%
  ggplot(aes(x = Embarked, 
             y = prop, 
             fill = fct_inseq(Pclass))) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of passengers in each class per port of embarkation",
       y = "Passengers [%]",
       x = "Place of embarkation",
       fill = "Class") +
  scale_fill_brewer(palette = "Dark2", type = "qual")
  
```


### Ticket
```{r}
train %>% 
  group_by(Ticket) %>% 
  filter(n() > 1) %>% 
  select(c(Ticket, Pclass, SibSp, Parch, Fare, Survived)) %>% 
  mutate(across(c(SibSp, Parch), as.numeric),
         WithFamily = SibSp + Parch) %>%
  ungroup() %>% 
  distinct(Ticket, .keep_all = TRUE) %>% 
```
```{r}
train %>% 
  group_by(Ticket) %>% 
  filter(n() > 1) %>% 
  ungroup() %>% 
  count(Pclass, Survived)
```

The ticket column contains numeric and mixed entries. We will investigate whether there are any differences in the ticket types, i.e. numeric vs. non-numeric, between different starting integers, different strings of the non-numeric, etc. First we will separate numeric and non-numeric types and group the ticket types based on strings and first integers.

```{r echo=TRUE, warning=FALSE}
train <- train %>%
  # transform values to numeric. if not possible, it will default to NA
  mutate(Numeric_ticket = !is.na(as.numeric(Ticket)),
         # transform to lower case to merge identical entries with different capitalization
         Ticket = tolower(Ticket),
         # create categorical values based on evaluation of numeric_ticket.
         Ticket_group = case_when(
           #  extract string if non-numeric
         Numeric_ticket == FALSE ~ str_extract(Ticket, "^[^\\s]+"),
         # extract first integer if numeric
         Numeric_ticket == TRUE ~ str_extract(Ticket, "^\\d")),
         Ticket_group = gsub("\\.", "", Ticket_group))
  
```

```{r echo=FALSE}
train %>% 
  count(Survived, Numeric_ticket, name = "Count") %>% 
  group_by(Numeric_ticket) %>% 
  mutate(Percentage = round(Count*100/sum(Count))) %>% 
  arrange(Numeric_ticket) 
```

```{r}
# group tickets by their first number or letter
train %>% 
  mutate(Ticket_class = str_extract(Ticket, "^\\S{1}")) %>% 
  group_by(Ticket_class)
  # count(Survived, Numeric_ticket, name = "Count") %>% 
  # group_by(Numeric_ticket) %>% 
  # mutate(Percentage = round(Count*100/sum(Count))) %>% 
  # arrange(Numeric_ticket) 
```


There is no difference in survival rate between numerical and non-numerical tickets. 
We can further divide each class.

```{r echo=FALSE}
train %>% 
  count(Survived, Ticket_group, name = "Count") %>% 
  group_by(Ticket_group) %>% 
  mutate(Percentage = round(Count*100/sum(Count))) %>%
  # sort groups based on highest number of survivors
  arrange(desc(Count * (Survived == 1)), .by_group = TRUE)
```

There are too many groups to infer any meaningful pattern. Some groups could probably be merged but without any detailed information we can only speculate whether "a/5" and "a5" are the same.

From "ticket" we can create a new variable "sharing_ticket", which indicates how many passengers shared the same ticket.
```{r}
train <- train %>% 
  mutate(Fam_size = SibSp + Parch) %>%
  group_by(Fam_size, Ticket) %>% 
  mutate(Sharing_ticket = n()) %>% 
  ungroup()
```


### EDA summary

Sex and passenger class are highly correlated with survival.

The cabin variable has too many missing values to directly impute them. 

Passenger ID and Ticket can also be dropped as they hold no predictive value.




```{r}
base_rec <- 
  base_rec %>%
  update_role(Ticket, PassengerId, Cabin, Name, new_role = "info_var")

summary(base_rec) %>% arrange(desc(role))
```

```{r}
age_pre <- ggplot(train, aes(Age)) +
  geom_density(na.rm = TRUE)

fare_pre <- 
  ggplot(train, aes(log(Fare))) +
  geom_density(na.rm = TRUE)
  
ggpubr::ggarrange(age_pre, fare_pre, ncol = 2, labels = )
```

```{r recipe}
#  create recipe
# Impute
# Handle factor levels
# Individual transformations for skewness and other issues
# Discretize (if needed and if you have no other choice)
# Create dummy variables
# Create interactions
# Normalization steps (center, scale, range, etc)
# Multivariate transformation (e.g. PCA, spatial sign, etc)

rec <- base_rec %>% 
  # imputation
  step_impute_mode(Embarked) %>% 
  step_impute_bag(Age)
  step_mutate(Fare = sub(0, NA, Fare)) %>% 
  step_impute_bag(Age) %>% 
  # factor levels
  step_other(SibSp, Parch, threshold = .05, other = "other") %>% 
  # transformation
  step_log(Fare) %>% 
  # Create dummy variables
  step_dummy(all_factor_predictors()) %>% 

  step

```



```{r recipe_titles}
rec <- recipe(Survived ~ ., train) %>% 
  # step_select(!c(starts_with("Ticket"), Passenger_ID, Cabin)) %>% 
  step_impute_mode(Embarked) %>% 
  step_mutate(Name = gsub(" ", "", Name),
             Title = str_extract(Name, ",([a-zA-Z]*)"),
             Title = gsub(",", "", Title)) %>% 
  step_select(-Name) %>% 
  step_impute_knn(Age, 
                  impute_with = imp_vars(Title, Ticket, SibSp),
                  neighbors = 3) 
```

```{r eval=FALSE}
titanic_rec <- rec %>% 
  
```

```{r}

```

### Imputation


#### Age 

For the imputation of age, extract the titles the name column. 
Then create title groups and calculate the mean and standard deviation for each and then use these to sample and impute from. 

```{r}
options(tibble.print_max = 17, tibble.print_min = 4)

train%>% 
    # Remove spaces
    mutate(Name = gsub(" ", "", Name),
           # Extract string following (including) comma
         Title = str_extract(Name, ",([a-zA-Z]*)"),
         # Remove comma
         Title = gsub(",", "", Title)) %>% 
  select(!Name) %>% 
  group_by(Title) %>% 
  count(Title, name = "Count", sort = TRUE)
```


```{r echo=TRUE}

extract_titles <- function(df, na.rm = FALSE){

  df <- df %>% 
      mutate(Name = gsub(" ", "", Name),
             Title = str_extract(Name, ",([a-zA-Z]*)"),
             Title = gsub(",", "", Title)) %>% 
    # remove name variable
      select(-Name)
  
  return(df)
}
```

```{r echo=TRUE}
train_with_titles <- extract_titles(train)
```

```{r echo=TRUE, paged.print=FALSE}
train_with_titles %>% 
  count(Title) %>% 
  arrange(desc(n))
```
We now have extracted 17 distinct titles of which most have only few observations. 
To perform robust imputation we need enough data in each group.
Therefore will will merge all titles into four groups: Mr, Miss, Mrs and Master. 
We will merge "Lady"/"Countess"/"Ms" with "Mrs", "Mme" with "Miss", and all other titles with "Mr".
Create another function called "aggregate titles".
```{r echo=TRUE}
aggregate_titles <- function(df, na.rm = FALSE){
  # create a negative selector helper function
  `%nin%` <- purrr::negate(`%in%`)
  
  df <- df %>% 
      mutate(
        # Aggregate female titles first
        Title = case_when(
          Title %in% c("Lady", "theCountess", "Ms") ~ "Mrs",
          Title == "Mme" ~ "Miss",
          TRUE ~ Title
          ),
        # Aggregate everything else
        Title = case_when(
          Title %nin% c("Mr", "Miss", "Mrs", "Master") ~ "Mr",
          TRUE ~ Title
          )
        )
  
  return(df)
}
```

```{r echo=T, results='hide'}
train_with_titles <- aggregate_titles(train_with_titles)
```

```{r echo=FALSE}
train_with_titles %>% 
  group_by(Title) %>% 
  summarise(Count  = n(), 
            Mean   = round(mean(Age, na.rm = TRUE), 1), 
            Median = median(Age, na.rm = TRUE), 
            SD     = round(sd(Age, na.rm = TRUE), 1)) %>% 
  kable("simple")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
train_with_titles %>% 
  group_by(Title) %>% 
  filter(!is.na(Age)) %>% 
  ggplot(aes(x = Title,
             y = Age)) +
  geom_boxjitter(aes(fill = Title),
                 position = position_dodge(width = 0.5),
                 width = 0.4,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4) +
  labs(title = "Age distribution across title groups",
       y = "Age",
       x = "Title") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual()
```

```{r echo=FALSE}
train_with_titles %>% 
  group_by(Title) %>% 
  filter(!is.na(Age)) %>%
  ggplot(aes(x = Age)) +
  geom_density(aes(color = Title)) +
  facet_grid(Title ~ .) +
  labs(title = "Density plot of age",
       y = "Dnesity",
       x = "Age",
       color = "Title") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_color_manual()

```


Since the title "Miss" is used for unmarried women, we will split the "Miss" group based on whether they were traveling with Parents and/or children -  assuming that those without are older.  
```{r echo=FALSE, message=FALSE, warning=FALSE}
train_with_titles %>% 
  filter(!is.na(Age)) %>%  # & Title == "Miss") %>% 
  mutate(Family = case_when(
    Parch == 0 ~ "No",
    TRUE ~ "Yes"
  )) %>% 
  ggplot(aes(x = Family,
             y = Age)) +
  geom_boxjitter(aes(fill = Family),
                 position = position_dodge(width = 0.5),
                 width = 0.4,
                 jitter.shape = 21,
                 jitter.color = NA,
                 outlier.shape = 1,
                 errorbar.draw = TRUE,
                 errorbar.length = 0.4) +
  labs(title = "Age distribution of passengers with family",
       y = "Age",
       x = "Title") +
  facet_grid(~Title) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual()
```
Our assumption turns out to be true so we will consider it within our imputation function.
The same assumption seems to be true for young men ("Master") but with just one observation there is not enough data to support it.
Arguably one could also merge "Mr" and "Mrs" as their distributions are very similar. 

```{r echo=TRUE}
impute_age <- function(df){
  # set seed for reproducibility
  set.seed(1234)
  
  # helper function to calculate mean+sd
  group_mu_sd <- df %>% 
    mutate(Title = ifelse(Title == "Miss" & "Parch" > 0, "Miss_fam", Title)) %>% 
    group_by(Title) %>% 
    summarise(Mean = round(mean(Age, na.rm = TRUE)),
              SD = round(sd(Age, na.rm = TRUE)),
              .groups = "drop")
  
  # helper function for sampling from a normal distribution but within one standard deviation from the mean
  sample_rnorm_within_bounds <- function(mean, sd, lower = mean - sd, upper = mean + sd, max_attempts = 1000) {
    for (i in 1:max_attempts) {
      sample <- rnorm(1, mean, sd)
      if (sample >= lower && sample <= upper) {
        return(sample)
      }
    }
    stop(paste("Failed to sample a value within bounds after", max_attempts, "attempts."))
  }
  
  # impute age by random sampling from normal distribution
  df_age_imputed <- df %>%
    mutate(Title = ifelse(Title == "Miss" & "Parch" > 0, "Miss_fam", Title)) %>%
    left_join(group_mu_sd, by = "Title") %>% 
    group_by(Title) %>% 
    rowwise() %>% 
    mutate(Age = ifelse(
      is.na(Age),
      round(
        sample_rnorm_within_bounds(mean = Mean, sd = SD)
        ),
      Age
      )
    ) %>% 
    ungroup() %>% 
    select(-c(Mean, SD))
  
  return(df_age_imputed)
}
```

Again we will compare the number of missing values before 
```{r echo=FALSE}
train_with_titles %>% 
  group_by(Title) %>% 
  summarise(Missing = sum(is.na(Age))) %>% 
  kable("simple")
```

...and after imputation
```{r echo=FALSE}
impute_age(train_with_titles) %>% 
  group_by(Title) %>% 
  summarise(Missing = sum(is.na(Age)),
            .groups = "drop") %>% 
  kable("simple")
```

Let us compare the mean and standard deviation of the original age values
```{r echo=FALSE}
train_with_titles %>% 
  filter(!is.na(Age)) %>%
  mutate(Title = ifelse(Title == "Miss" & "Parch" > 0, "Miss_fam", Title)) %>% 
  group_by(Title) %>% 
    summarise(Count = n(),
      Mean = round(mean(Age, na.rm = TRUE), 1),
      SD   = round(sd(Age, na.rm = TRUE), 1)) %>% 
  kable("simple")
```

...with the imputed values
```{r echo=FALSE}
impute_age(train_with_titles) %>% 
  group_by(Title) %>% 
    summarise(Count = n(),
      Mean = round(mean(Age, na.rm = TRUE), 1),
      SD   = round(sd(Age, na.rm = TRUE), 1),
      .groups = "drop") %>% 
  kable("simple")
```

We can also compare the densities of Age before and after imputation.
```{r echo=FALSE}
train_with_titles %>% 
  filter(!is.na(Age)) %>% 
  ggplot() +
  geom_density(aes(x = Age)) +
  labs(title = "Density plot of age before imputation",
       y = "Density",
       x = "Age") +
  theme_minimal()
```

```{r echo=FALSE}
impute_age(train_with_titles) %>% 
  ggplot() +
  geom_density(aes(x = Age)) +
    labs(title = "Density plot of age after imputation",
       y = "Density",
       x = "Age") +
  theme_minimal()

```

```{r}
train_imputed <- impute_age(train_with_titles) %>% 
  select(-Title)
```

### Cabin
The cabin variables has the highest proportion of missing values - too many to impute them. 
However, we will explore it and try to extract meaningful information.
The values start with a character (the deck) followed by the cabin number. We will group cabin values based on the deck.

```{r}
train_imputed <- train_imputed %>% 
  mutate(Deck = substring(Cabin, 1, 1),
         # rename NAs as "M" for missing
         Deck = replace_na(Deck, "M"))

train_imputed %>% 
  count(Deck, name = "Observations") %>% 
  kable()
```

Let us look at the distributions of survivors and passenger classes across decks.
```{r Deck Survived, echo=FALSE}
train_imputed %>% 
    # filter(!is.na(Embarked)) %>%
  group_by(Deck, Survived) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(prop = count/sum(count),
         Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>%
  ggplot(aes(x = Deck, 
             y = prop, 
             fill = Survived)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of survivors on each deck",
       y = "Passengers [%]",
       x = "Deck") +
  theme_minimal() +
  scale_fill_manual()
```


On https://titanicfacts.net/titanic-ship/ we find some information about the passenger decks: 
''''
How many decks did the Titanic have?

There were 10 decks in total. From top to bottom they were the Boat Deck, the Promenade Deck (deck A), passenger decks B to G, Orlop Deck, and the Tank Top.
'''

Let us investigate how the individual classes are distributed across the different decks.
```{r Deck PClass, echo=FALSE}
train_imputed %>% 
    # filter(!is.na(Embarked)) %>%
  group_by(Deck, Pclass) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(prop = count/sum(count)) %>%
  ggplot(aes(x = Deck, 
             y = prop, 
             fill = Pclass)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of passengers from each class on each deck",
       y = "Passengers [%]",
       x = "Deck",
       fill = "Class") +
  theme_minimal() +
  scale_fill_manual()
```
We already know that passenger class and fare are correlated with survival. 
Let's see if there are any differences in survival rates within each class based on the deck of the cabin.
We will exclude decks A, T and missing values.
```{r}
train_imputed %>% 
  count(Deck, Pclass, Survived) %>% 
  mutate(Survived = case_match(Survived,
                              "0" ~ "deceased",
                              "1" ~ "survived")) %>% 
  group_by(Deck, Pclass) %>% 
  mutate(Passengers = sum(n)) %>% 
  group_by(Survived) %>% 
  mutate(Survived_prop = n*100/Passengers) %>% 
  filter(Deck %in% c("B", "C", "D", "E", "F", "G")) %>% 
  ungroup() %>% 
  ggplot(aes(x = interaction(Deck, Pclass),
             y = Survived_prop, 
             fill = Survived,
             label = n
             )) +
  geom_bar(stat = "identity", position = "fill") +
  geom_label(aes(label = n),
             position = position_stack(vjust = 0.5),
             size = 3,
             show.legend = FALSE) +
  labs(title = "Survival rates in each class based on cabin deck",
       y = "Passengers [%]",
       x = "Deck, Class") +
  theme_minimal() +
  scale_fill_manual()
  
```


```{r Fares in each Class/Deck, echo=FALSE, message=FALSE, warning=FALSE}
train_imputed %>% 
  group_by(Deck, Pclass) %>% 
  ggplot(aes(x = interaction(Deck, Pclass), 
             y = Fare)) +
  geom_boxplot(aes(fill = Pclass)) +
  labs(title = "Distribution of fares for each deck and passenger class",
       y = "Fare",
       x = "Deck, Class") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_manual()
```
One could argue that fare decreases as we get to the lower decks but this is only true for the first class - where sample size is much higher.
We could thus assign missing values to decks based on fare for 1st class-passengers. However, for second and third-class passengers, based on lacking data, we simply cannot make this assumption. With more data, we could bin the fares based on the quantiles of the distribution of fares in each respective class.

Based on the information above we will merge decks A and T with deck E to match the original passenger decks.
```{r echo=TRUE}
train_imputed <- train_imputed %>% 
  mutate(Deck = case_when(
    Deck %in% c("A", "T") ~ "E",
    .default = Deck
  ))
```

Passengers from the first class populated decks B to E, second class decks D to F and third class decks E to G. Thus we will create 4 bins for class 1 and 3 bins for classes 2 and 3, respectively.
First, we need to find the respective quantiles.
```{r}
Deck_qunatiles <- train_imputed %>% 
  filter(Deck == "M") %>% 
  group_by(Pclass) %>% 
  summarise(
    Q1 = if_else(Pclass == 1, quantile(Fare, probs = 0.25), quantile(Fare, probs = 1/3)),
    Q2 = if_else(Pclass == 1, round(quantile(Fare, probs = 0.5)), round(quantile(Fare, probs = 2/3))),
    Q3 = if_else(Pclass == 1, quantile(Fare, probs = 0.75), NA_real_),
    .groups = "drop"
      ) %>% 
  distinct()

Deck_qunatiles
```

```{r}
train_imputed %>% 
  
  
```


### Feature encoding and scaling

In our approach we will compare different classification techniques.
While tree-based models are insensitive to different feature scales and outliers, distance-based models like k-nearest neighbors will perform better when features are scaled uniformly.
There are still two numeric variables in the data: fare and age. 
Let us again look at their distributions.

```{r echo=FALSE, message=FALSE, warning=FALSE}
numeric_train <- train_fe %>% 
  select(Age, Fare)%>% 
  names()

num_plots <- list()
colors_num <- lette("GrandBudapest1", length(numeric_train), type = "discrete")

for (i in seq_along(numeric_train)) {
  
  var <- numeric_train[i]
  
  p <- train_fe %>% 
    ggplot(aes_string(x = var)) +
    geom_histogram(fill = colors_num[[i]],
                   color = "black") +
    labs(title = paste("Distribution of", var),
       y = "Count",
       x = var) +
    theme_minimal() +
    theme(legend.position = "none")
  
  num_plots <- append(num_plots, list(p))
}

do.call(grid.arrange, c(num_plots, ncol = 2))
```

```{r}
numeric_train <- train_fe %>% 
  select(Age, Fare)%>% 
  names()

num_plots <- list()
colors_num <- lette("GrandBudapest1", length(numeric_train), type = "discrete")

for (i in seq_along(numeric_train)) {
  
  var <- numeric_train[i]
  
  p <- train_fe %>% 
    ggplot(aes_string(x = var)) +
    geom_boxplot(fill = colors_num[[i]],
                 outlier.colour =  lette("GrandBudapest1")[3]) +
    labs(title = paste("Boxplot of", var),
       x = var) +
    coord_flip() +
    theme_minimal() +
    theme(axis.text.y = element_blank()) +
    theme(legend.position = "none")
  
  num_plots <- append(num_plots, list(p))
}

do.call(grid.arrange, c(num_plots, ncol = 2))
```
We want to scale both age and fare to [0, 1] in order to have a uniform feature scaling.
This is achieved by Min-Max scaling. However, Min-Max scaling is sensitive to outliers - of which we find many in fare.
Therefore we will first apply log-transform in order to decrease the variance.
```{r}
train_fe <- train_fe %>%
  # log1p adds 1 to each value in order to avoid -Inf values after transformation
  mutate(Fare = log1p(Fare))
```

Now we will scale both variables to [0, 1].
```{r}
preproc_scale <- preProcess(train_fe[, c("Fare", "Age")], method = "range", rangeBounds = c(0, 1))
data_scaled <- predict(preproc_scale, newdata = train_fe[, c("Fare", "Age")])
```

```{r message=FALSE, warning=FALSE}
data_scaled %>% 
  ggplot() +
  geom_histogram(aes(x=Fare),
                 fill = lette("GrandBudapest1")[2],
                 color = "black") +
  labs(title = "Distribution of fare after log-transformation and scaling",
       y = "Count",
       x = "Fare") +
  theme_minimal()
  
```
Fare is now distributed across the whole interval.

```{r}
summary(train_fe)
```


For ease of application we will implement all of the steps outlined above into a handy function for data clean up and imputation.
```{r}
clean_and_impute <- function(df, na.rm = FALSE){
  
  # Helper functions----
  # create a negative selector helper function
  `%nin%` <- purrr::negate(`%in%`)
  
  # helper function for sampling from a normal distribution within one standard deviation from the mean
  sample_rnorm_within_bounds <- function(mean, sd, lower = mean - sd, upper = mean + sd, max_attempts = 1000) {
    for (i in 1:max_attempts) {
      sample <- rnorm(1, mean, sd)
      if (sample >= lower && sample <= upper) {
        return(sample)
      }
    }
    stop(paste("Failed to sample a value within bounds after", max_attempts, "attempts."))
  }
  
  # remove columns
  df_clean <- df %>% 
    select(-c(Cabin, PassengerId, Ticket))
  
  # Imputation----
  
  # Impute location of embarkation
  df_emb <- df_clean %>% 
    mutate(Embarked = case_when(
      is.na(Embarked) ~ impute_mode(train_cln, "Embarked"),
      TRUE ~ as.character(Embarked)
    ))
  
  # Impute_fare 
  grouped_median_fares <- df_emb %>% 
    group_by(Pclass, Parch) %>% 
    summarise(Fare_median = median(Fare, na.rm = TRUE))
  
  df_fare <- df_clean %>%
    left_join(grouped_median_fares, by = "Pclass") %>% 
    mutate(Fare = coalesce(Fare, Fare_median)) %>%
    select(-Fare_median)
  
  # Check for complete values
  ifelse(sum(is.na(df_fare$Fare)) == 0, 
         print("Fare successfully imputed!"), 
         print("Something went wrong!"))
  
  # Impute age 
  # Create title variable
  df_titles <- df_fare %>% 
    mutate(Name = gsub(" ", "", Name),
           Title = str_extract(Name, ",([a-zA-Z]*)"),
           Title = gsub(",", "", Title)) %>% 
  # remove name variable
    select(!Name) %>% 
    mutate(
      # Aggregate female titles first
      Title = case_when(
        Title %in% c("Lady", "theCountess", "Ms") ~ "Mrs",
        Title == "Mme" ~ "Miss",
        TRUE ~ Title
        ),
      # Aggregate everything else
      Title = case_when(
        Title %nin% c("Mr", "Miss", "Mrs", "Master") ~ "Mr",
        TRUE ~ Title
        )
      ) 

  # Helper function to calculate mean+sd of each title group
  group_mu_sd <- df_titles %>% 
    mutate(Title = ifelse(Title == "Miss" & Parch > 0, "Miss_fam", Title)) %>% 
    group_by(Title) %>% 
    summarise(Mean = round(mean(Age, na.rm = TRUE)),
              SD = round(sd(Age, na.rm = TRUE)))
  
  # impute age by random sampling from normal distribution
  df_imputed <- df_titles %>%
    mutate(Title = ifelse(Title == "Miss" & Parch > 0, "Miss_fam", Title)) %>%
    left_join(group_mu_sd, by = "Title") %>% 
    group_by(Title) %>% 
    rowwise() %>% 
    mutate(Age = ifelse(
      is.na(Age),
      round(rnorm(1, mean = Mean, sd = SD)),
      Age
    )) %>% 
    ungroup() %>% 
    select(-c(Mean, SD, Title))
  
  # Check for complete values
  ifelse(sum(is.na(df_imputed$Age)) == 0, 
         print("Fare successfully imputed!"), 
         print("Something went wrong!"))
  
  return(df_imputed)
  
}
```



## Modeling

```{r}
# specify models
lg_spec <- logistic_reg(penalty = tune(), mixture = 1) %>% set_engine("glm")

rf_spec <- rand_forest(trees = 100) %>% set_engine("ranger") %>% set_mode("classification")

svm_spec <- svm_rbf() %>% set_mode("classification")
```

```{r}
# add model to recipe
doParallel::registerDoParallel()

mod_metrics <- metric_set(roc_auc, precision, mcc, f_meas)

glm_rs <- training_flow %>% 
  add_model(glm_spec) %>% 
  fit_resamples(
    resamples = train_folds,
    metrics = mod_metrics,
    control = control_resamples(save_pred = T)
  )
```

